Suitability prevalence area index in late quaternary explains genetic diversity in Tassel eared Squirrels

### 2023-07-29 manuscript_english.Rmd
finish abstract

## Abstract
As a case study we consider squirrels (*Sciurus aberti*), a species currently distributed in disjunct patches from the southern Rocky Mountains in the United States to the northern Sierra Madre Occidental in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution [@bono2018genome; @burgin2018many].

Our results reveal that fitness prevalence corresponds to the fixation index of *S. aberti* populations with respect to a source population. Populations closer to the historical endemic area present a higher genetic diversity and a lower $F_{st}$ value. Finally, this study allows us to add a biogeographic explanation to the results obtained with population genetic methods and to generate maps of this structure as tools to support conservation with a perspective that integrates both population genetics and historical patterns of species distribution.

@@

We define $\vec{e}_i =(e_1, e_2, ..., e_i )$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables. Then $R_i(\vec{e}_i)$ is a function that relates fitness to environment,
that is, for a species with non-overlapping generations, the population's net reproductive rate $R_i$ is a measure of fitness that depends on environmental conditions.

@@

This equation was proposed by Soberon and Peterson @soberon2020shape to relate demographics to a function of fitness and environment. In that study they tested the validity with monthly variables and empirical values on a short time scale. relating life history parameters and the environmental suitability of niche models.

@@

Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and that for each trajectory the fitness function changes and therefore the suitability function changes but not accordingly. in a homogeneous way, if not more so that each site presents different displacement rates and different demographic rates.

The objective of this theoretical approach is not to know the shape of the function of $R_0$ but rather to find this local variation relating this value of $R_0$ and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).

Es decir:

@@

We call this harmonic mean of environmental suitability *Suitability prevalence index *($S$) given between $0$ and $1$, since it reflects the rate of population growth as a function of environment over a period of time. Thus we have a way to check that the loss of heterozygosity (through the fixation index $F_{st}$) is inversely proportional to the fitness prevalence for an $i$ site in the geography:

$$
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
$$
If we consider all sites in the available geography, we call this harmonic mean of fitness *Suitability prevalence Area* $\mathbf{S} = \{S_i | S_i > k\}$ since we expect that areas where the $F_{st}$ is lower are in areas where the prevalence over time is higher, i.e. environmental conditions remain suitable for a significant period of time and therefore no effects of genetic drift of populations have occurred. Conversely, it is expected that in populations with a high fixation index value, fitness will not have prevailed over the same time interval even though conditions currently exist and populations are present.

@@

To test our model we use the Aberti squirrel (*Sciurus aberti*) as a case study because:

- Since it is a medium-sized mammal (0.5 kg) its identification is easy.
- The subspecies are geographically isolated, since they are found in pine forests surrounded by desert. Therefore we can consider the assumption of isolation by distance [@wright1943isolation]. 
- There are sufficient occurrence records for the entire range of the known distribution.
- It has low speciation rate (0.25) [@upham2019inferring], therefore, we assume that its fundamental niche is stable over time.
- There is fixation index information [@bono2018genome] for different populations of the subspecies, which is useful for using SPA as a way to explain genetic differentiation between populations.

@@

To test our hypothesis of genetic diversity as a function of environmental suitability prevalence, from a set of genetically structured populations, a matrix of fixation indices between populations is considered. For this work we considerer the data from @@bono2018genome. It consideres 10 populations along the known distribution and provides values of $F_st$ between this populations in matrix form. The matrix was transformed into a table as pairs between populations with index values between each pair. The values of the population with itself, i.e. $0$, were included. Subsequently, georeferences were obtained for each population considered in the fixation index matrix. From this geo-referencing the SPA value for each population was extracted. Finally, the SPA value was aligned in the table of fixation index pairs. In total the table has 4 columns: source population, target population, fixation index and SPA of the target population. This table was separated by source population groups to generate a set of linear models between SPA values and fixation indexes $F_{st}$. An analysis was performed for each source population with the hypothesis that the correlation between SPA and fixation index exists and is non-zero if a dispersal process has occurred from the historical endemic area of the taxon.

@@

For all the linear regressions performed (Table 1), the only one that had statistical significance is the one for the Sciurus aberti barberi population with an R-squared value of 0.794.

```{r, echo=FALSE}
aberti_pops <- read.csv(file = "all_tables/fst_SPA_rsquared.csv")

knitr::kable(aberti_pops, "simple", align = "ccc", caption = "R squared from linear models (fixation index as a function of suitability prevalence area). The only R squared with statistical significance is barberi")
```

>>>>

## Abstract

Genetic variation among populations exhibits biogeographic patterns. In this paper we establish a theoretical framework based on environmental Grinellian niches as a tool for modeling species distributions, as well as their projected past environmental scenarios to relate the present evolutionary history in both the geographic distribution and genetic structure of their populations. Thanks to the advance of next-generation sequencing technologies, as well as more accurate climate models and an enormous amount of information stored in biological collections, it is possible to implement this theoretical framework with the help of the joint analysis of spatial, environmental and genetic data, with tools that are widely distributed among biogeographers and evolutionary biologists. Our results reveal that when we have case of isolation by distance, suitability prevalence along time corresponds to the fixation index of populations with respect to a source population. Populations closer to the historical endemic area present a higher genetic diversity and a lower $F_{st}$ value. This framework allows us to add a biogeographic explanation to the results obtained with population genetic methods and to generate maps of this structure as tools to support conservation with a perspective that integrates both population genetics and historical patterns of species distribution.

@@

We define $\vec{e}_i =(e_1, e_2, ..., e_i )$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables. Then $R_i(\vec{e}_i)$ is a function that relates evolutionary fitness to environment,
that is, for a species with non-overlapping generations, the population's net reproductive rate $R_i$ is a measure of evolutionary fitness that depends on environmental conditions.

@@

This equation was proposed by Soberon and Peterson @soberon2020shape to relate demographics to a function of evolutionary fitness and environment. In that study they tested the validity with monthly variables and empirical values on a short time scale. relating life history parameters and the environmental suitability of niche models.

@@

Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and that for each trajectory, the fitness function changes and, therefore, the suitability function changes but not accordingly, in a homogeneous way, if not more so that each site presents different displacement rates and different demographic rates.

The goal of this theoretical approach is not to know the shape of the function of $R_0$ but rather to find this local variation relating this value of $R_0$ and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum), i.e. :

@@

We call this harmonic mean of environmental suitability *Suitability prevalence index *($S$) given between $0$ and $1$, since it reflects the rate of population growth as a function of environment over a period of time. Thus we have a way to check that the loss of heterozygosity (through the fixation index $F_{st}$) is inversely proportional to the suitability prevalence for an $i$ site in the geography:

$$
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
$$
If we consider all sites in the available geography, we call this harmonic mean *Suitability Prevalence Area* $\mathbf{S} = \{S_i | S_i > k\}$ (SPA) since we expect that areas where the $F_{st}$ is lower are in areas where the prevalence over time is higher, i.e. environmental conditions remain suitable for a significant period of time and therefore no effects of genetic drift of populations have occurred. Conversely, it is expected that in populations with a high fixation index value, suitability will not have prevailed over the same time interval even though conditions currently exist and populations are present.

@@

As a case study we consider squirrels (*Sciurus aberti*), a species currently distributed in disjunct patches from the southern Rocky Mountains in the United States to the northern Sierra Madre Occidental in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution [@bono2018genome; @burgin2018many]. In order to select this species, we list the following considerations:

- Since it is a medium-sized mammal (0.5 kg) its identification is easy.
- The subspecies are geographically isolated, since they are found in pine forests surrounded by desert. Therefore we can consider the assumption of isolation by distance [@wright1943isolation]. 
- It has low speciation rate (0.25) [@upham2019inferring], therefore, we assume that its fundamental niche is stable over time.
- There is fixation index information [@bono2018genome] for different populations of the subspecies, which is useful for using SPA as a way to explain genetic differentiation between populations.

@@

To test our hypothesis of genetic diversity as a function of environmental suitability prevalence, from a set of genetically structured populations, a matrix of fixation indices between populations is considered. For this work we considerer the data from @bono2018genome. It consideres 10 populations along the known distribution and provides values of $F_st$ between this populations in matrix form. The matrix was transformed into a table as pairs between populations with index values between each pair. The values of the population with itself, i.e. $0$, were included. Subsequently, georeferences were obtained for each population considered in the fixation index matrix. From this geo-referencing the SPA value for each population was extracted. Finally, the SPA value was aligned in the table of fixation index pairs. In total the table has 4 columns: source population, target population, fixation index and SPA of the target population. This table was separated by source population groups to generate a set of linear models between SPA values and fixation indexes $F_{st}$. An analysis was performed for each source population with the hypothesis that the correlation between SPA and fixation index exists and is non-zero if a dispersal process has occurred from the historical endemic area of the taxon.

@@

For all the linear regressions performed (Table 1), the only one that had statistical significance is the one for the Sciurus aberti barberi population with an $R^2$ value of $0.794$.

|          Population          | R squared | p value |
|:----------------------------:|:---------:|:-------:|
|   _S. a. aberti Carson-SFW_  |     0     |  0.962  |
| _S. a. aberti Coconino-Gila_ |   0.131   |  0.304  |
|    _S. a. aberti MT-Zuni_    |   0.002   |  0.894  |
|    _S. a. aberti San Juan_   |     0     |  0.952  |
|        _S. a. barberi_       |   0.794   |  0.001  |
|     _S. a. chuscensis E_     |   0.002   |  0.892  |
|     _S. a. chuscensis W_     |   0.018   |  0.713  |
|   _S. a. ferreus Carson E_   |    0.04   |  0.581  |
|     _S. a. ferreus Pike_     |     0     |  0.967  |
|      _S. a. kaibabensis_     |   0.005   |  0.841  |

### 2023-07-29 manuscript_english.tex
finish abstract

As a case study we consider squirrels (\emph{Sciurus aberti}), a species
currently distributed in disjunct patches from the southern Rocky
Mountains in the United States to the northern Sierra Madre Occidental
in Mexico for which it is possible to find reliable information on both
the genetic structure of its populations and its current distribution
(Bono et al. 2018; Burgin et al. 2018).

Our results reveal that fitness prevalence corresponds to the fixation
index of \emph{S. aberti} populations with respect to a source
population. Populations closer to the historical endemic area present a
higher genetic diversity and a lower \(F_{st}\) value. Finally, this
study allows us to add a biogeographic explanation to the results
obtained with population genetic methods and to generate maps of this
structure as tools to support conservation with a perspective that
integrates both population genetics and historical patterns of species
distribution.

@@

We define \(\vec{e}_i =(e_1, e_2, ..., e_i )\) as a vector of
environmental variables, that is, a point in the space of \(v\)
environmental variables. Then \(R_i(\vec{e}_i)\) is a function that
relates fitness to environment, that is, for a species with
non-overlapping generations, the population's net reproductive rate
\(R_i\) is a measure of fitness that depends on environmental
conditions.

@@

This equation was proposed by Soberon and Peterson Soberón and Peterson
(2020) to relate demographics to a function of fitness and environment.
In that study they tested the validity with monthly variables and
empirical values on a short time scale. relating life history parameters
and the environmental suitability of niche models.

@@

Once we have an approximation of this volume, we consider that each
point in the geography is static and has a trajectory in the
environmental space and that for each trajectory the fitness function
changes and therefore the suitability function changes but not
accordingly. in a homogeneous way, if not more so that each site
presents different displacement rates and different demographic rates.

The objective of this theoretical approach is not to know the shape of
the function of \(R_0\) but rather to find this local variation relating
this value of \(R_0\) and the function \(\mathbf{N}_F\) with a function
\$ S\_i(~vec\{e\}\_i)\$ of environmental suitability that can be
estimated from known methods of applied mathematics and machine learning
(for example the case of Maxent, Phillips, Anderson, and Schapire
(2006)).

Es decir:

@@

We call this harmonic mean of environmental suitability
\emph{Suitability prevalence index }(\(S\)) given between \(0\) and
\(1\), since it reflects the rate of population growth as a function of
environment over a period of time. Thus we have a way to check that the
loss of heterozygosity (through the fixation index \(F_{st}\)) is
inversely proportional to the fitness prevalence for an \(i\) site in
the geography:

\[
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
\] If we consider all sites in the available geography, we call this
harmonic mean of fitness \emph{Suitability prevalence Area}
\(\mathbf{S} = \{S_i | S_i > k\}\) since we expect that areas where the
\(F_{st}\) is lower are in areas where the prevalence over time is
higher, i.e.~environmental conditions remain suitable for a significant
period of time and therefore no effects of genetic drift of populations
have occurred. Conversely, it is expected that in populations with a
high fixation index value, fitness will not have prevailed over the same
time interval even though conditions currently exist and populations are
present.

@@

To test our model we use the Aberti squirrel (\emph{Sciurus aberti}) as
a case study because:

\begin{itemize}
\tightlist
\item
  Since it is a medium-sized mammal (0.5 kg) its identification is easy.
\item
  The subspecies are geographically isolated, since they are found in
  pine forests surrounded by desert. Therefore we can consider the
  assumption of isolation by distance (Wright 1943).
\item
  There are sufficient occurrence records for the entire range of the
  known distribution.
\item
  It has low speciation rate (0.25) (Upham, Esselstyn, and Jetz 2019),
  therefore, we assume that its fundamental niche is stable over time.
\item
  There is fixation index information (Bono et al. 2018) for different
  populations of the subspecies, which is useful for using SPA as a way
  to explain genetic differentiation between populations.
\end{itemize}

@@

To test our hypothesis of genetic diversity as a function of
environmental suitability prevalence, from a set of genetically
structured populations, a matrix of fixation indices between populations
is considered. For this work we considerer the data from @Bono et al.
(2018). It consideres 10 populations along the known distribution and
provides values of \(F_st\) between this populations in matrix form. The
matrix was transformed into a table as pairs between populations with
index values between each pair. The values of the population with
itself, i.e.~\(0\), were included. Subsequently, georeferences were
obtained for each population considered in the fixation index matrix.
From this geo-referencing the SPA value for each population was
extracted. Finally, the SPA value was aligned in the table of fixation
index pairs. In total the table has 4 columns: source population, target
population, fixation index and SPA of the target population. This table
was separated by source population groups to generate a set of linear
models between SPA values and fixation indexes \(F_{st}\). An analysis
was performed for each source population with the hypothesis that the
correlation between SPA and fixation index exists and is non-zero if a
dispersal process has occurred from the historical endemic area of the
taxon.

@@

For all the linear regressions performed (Table 1), the only one that
had statistical significance is the one for the Sciurus aberti barberi
population with an R-squared value of 0.794.

\begin{longtable}[]{@{}ccc@{}}
\caption{R squared from linear models (fixation index as a function of
suitability prevalence area). The only R squared with statistical
significance is barberi}\tabularnewline
\toprule\noalign{}
population & r.squared & p.value \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
population & r.squared & p.value \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
S. a. aberti Carson-SFW & 0.000 & 0.962 \\
S. a. aberti Coconino-Gila & 0.131 & 0.304 \\
S. a. aberti MT-Zuni & 0.002 & 0.894 \\
S. a. aberti San Juan & 0.000 & 0.952 \\
S. a. barberi & 0.794 & 0.001 \\
S. a. chuscensis E & 0.002 & 0.892 \\
S. a. chuscensis W & 0.018 & 0.713 \\
S. a. ferreus Carson E & 0.040 & 0.581 \\
S. a. ferreus Pike & 0.000 & 0.967 \\
S. a. kaibabensis & 0.005 & 0.841 \\
\end{longtable}

>>>>

Genetic variation among populations exhibits biogeographic patterns. In
this paper we establish a theoretical framework based on environmental
Grinellian niches as a tool for modeling species distributions, as well
as their projected past environmental scenarios to relate the present
evolutionary history in both the geographic distribution and genetic
structure of their populations. Thanks to the advance of next-generation
sequencing technologies, as well as more accurate climate models and an
enormous amount of information stored in biological collections, it is
possible to implement this theoretical framework with the help of the
joint analysis of spatial, environmental and genetic data, with tools
that are widely distributed among biogeographers and evolutionary
biologists. Our results reveal that when we have case of isolation by
distance, suitability prevalence along time corresponds to the fixation
index of populations with respect to a source population. Populations
closer to the historical endemic area present a higher genetic diversity
and a lower \(F_{st}\) value. This framework allows us to add a
biogeographic explanation to the results obtained with population
genetic methods and to generate maps of this structure as tools to
support conservation with a perspective that integrates both population
genetics and historical patterns of species distribution.

@@

We define \(\vec{e}_i =(e_1, e_2, ..., e_i )\) as a vector of
environmental variables, that is, a point in the space of \(v\)
environmental variables. Then \(R_i(\vec{e}_i)\) is a function that
relates evolutionary fitness to environment, that is, for a species with
non-overlapping generations, the population's net reproductive rate
\(R_i\) is a measure of evolutionary fitness that depends on
environmental conditions.

@@

This equation was proposed by Soberon and Peterson Soberón and Peterson
(2020) to relate demographics to a function of evolutionary fitness and
environment. In that study they tested the validity with monthly
variables and empirical values on a short time scale. relating life
history parameters and the environmental suitability of niche models.

@@

Once we have an approximation of this volume, we consider that each
point in the geography is static and has a trajectory in the
environmental space and that for each trajectory, the fitness function
changes and, therefore, the suitability function changes but not
accordingly, in a homogeneous way, if not more so that each site
presents different displacement rates and different demographic rates.

The goal of this theoretical approach is not to know the shape of the
function of \(R_0\) but rather to find this local variation relating
this value of \(R_0\) and the function \(\mathbf{N}_F\) with a function
\$ S\_i(~vec\{e\}\_i)\$ of environmental suitability that can be
estimated from known methods of applied mathematics and machine learning
(for example the case of Maxent, Phillips, Anderson, and Schapire
(2006)), i.e.~:

@@

We call this harmonic mean of environmental suitability
\emph{Suitability prevalence index }(\(S\)) given between \(0\) and
\(1\), since it reflects the rate of population growth as a function of
environment over a period of time. Thus we have a way to check that the
loss of heterozygosity (through the fixation index \(F_{st}\)) is
inversely proportional to the suitability prevalence for an \(i\) site
in the geography:

\[
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
\] If we consider all sites in the available geography, we call this
harmonic mean \emph{Suitability Prevalence Area}
\(\mathbf{S} = \{S_i | S_i > k\}\) (SPA) since we expect that areas
where the \(F_{st}\) is lower are in areas where the prevalence over
time is higher, i.e.~environmental conditions remain suitable for a
significant period of time and therefore no effects of genetic drift of
populations have occurred. Conversely, it is expected that in
populations with a high fixation index value, suitability will not have
prevailed over the same time interval even though conditions currently
exist and populations are present.

@@

As a case study we consider squirrels (\emph{Sciurus aberti}), a species
currently distributed in disjunct patches from the southern Rocky
Mountains in the United States to the northern Sierra Madre Occidental
in Mexico for which it is possible to find reliable information on both
the genetic structure of its populations and its current distribution
(Bono et al. 2018; Burgin et al. 2018). In order to select this species,
we list the following considerations:

\begin{itemize}
\tightlist
\item
  Since it is a medium-sized mammal (0.5 kg) its identification is easy.
\item
  The subspecies are geographically isolated, since they are found in
  pine forests surrounded by desert. Therefore we can consider the
  assumption of isolation by distance (Wright 1943).
\item
  It has low speciation rate (0.25) (Upham, Esselstyn, and Jetz 2019),
  therefore, we assume that its fundamental niche is stable over time.
\item
  There is fixation index information (Bono et al. 2018) for different
  populations of the subspecies, which is useful for using SPA as a way
  to explain genetic differentiation between populations.
\end{itemize}

@@

To test our hypothesis of genetic diversity as a function of
environmental suitability prevalence, from a set of genetically
structured populations, a matrix of fixation indices between populations
is considered. For this work we considerer the data from Bono et al.
(2018). It consideres 10 populations along the known distribution and
provides values of \(F_st\) between this populations in matrix form. The
matrix was transformed into a table as pairs between populations with
index values between each pair. The values of the population with
itself, i.e.~\(0\), were included. Subsequently, georeferences were
obtained for each population considered in the fixation index matrix.
From this geo-referencing the SPA value for each population was
extracted. Finally, the SPA value was aligned in the table of fixation
index pairs. In total the table has 4 columns: source population, target
population, fixation index and SPA of the target population. This table
was separated by source population groups to generate a set of linear
models between SPA values and fixation indexes \(F_{st}\). An analysis
was performed for each source population with the hypothesis that the
correlation between SPA and fixation index exists and is non-zero if a
dispersal process has occurred from the historical endemic area of the
taxon.

@@

For all the linear regressions performed (Table 1), the only one that
had statistical significance is the one for the Sciurus aberti barberi
population with an \(R^2\) value of \(0.794\).

\begin{longtable}[]{@{}ccc@{}}
\toprule\noalign{}
Population & R squared & p value \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\emph{S. a. aberti Carson-SFW} & 0 & 0.962 \\
\emph{S. a. aberti Coconino-Gila} & 0.131 & 0.304 \\
\emph{S. a. aberti MT-Zuni} & 0.002 & 0.894 \\
\emph{S. a. aberti San Juan} & 0 & 0.952 \\
\emph{S. a. barberi} & 0.794 & 0.001 \\
\emph{S. a. chuscensis E} & 0.002 & 0.892 \\
\emph{S. a. chuscensis W} & 0.018 & 0.713 \\
\emph{S. a. ferreus Carson E} & 0.04 & 0.581 \\
\emph{S. a. ferreus Pike} & 0 & 0.967 \\
\emph{S. a. kaibabensis} & 0.005 & 0.841 \\
\end{longtable}
